Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2020-03-02 manuscript/climate_sensitivity_draft.Rmd
#80

### Summary
- As climate change is driving increased drought in many forested regions around the world, mechanistic understanding of factors conferring drought resistance in trees is increasingly important. The dendrochronological record provides a window through which we can understand how tree size and species’ traits shape tree growth responses during droughts.

- We analyzed tree-ring records for twelve species of an oak-hickory forest - representing 97% of woody productivity - in the 25.6-ha ForestGEO plot in Virginia (USA) to determine how tree size, microenvironment characteristics, and species' traits shaped drought responses across the three strongest regional droughts over a 60-year period (1950 - 2009).

@@

- We conclude that hydraulic traits and tree height influence growth responses during drought, and can explain variation in the tree ring record spanning historical droughts. Thus, these factors can be useful for predicting future drought responses under climate change.

@@

Forests globally play a critical role in climate regulation [@bonan_forests_2008], yet there remains enormous uncertainty as to how the terrestrial carbon (C) sink, which is dominated by forests, will respond to climate change [@friedlingstein_climatecarbon_2006]. An important aspect of this uncertainty lies with physiological responses of trees to drought [@kennedy_implementing_2019]. In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014]; [Dai et al. 2018](https://link.springer.com/article/10.1007/s40641-018-0101-6)), often despite increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015]. Droughts, exasperated by climate change, have been affecting forests worldwide [@allen_global_2010], and are expected to continue as one of the most important drivers of forest change in the future [@allen_underestimation_2015]. Understanding and modeling forest responses to drought requires elucidation of how tree size, microenvironment, and species’ traits jointly influence individual-level drought resistance, and the extent to which their influence is consistent across droughts. However, it has proved difficult to resolve factors affecting tree growth during drought with available forest census data, which only rarely captures extreme drought, and with tree-ring records, which rarely represent many species and size classes at a single site and are not necessarily associated with ecological data. The aim of this study was to pair tree-ring and forest plot data to test how tree size and traits shaped growth responses to historical droughts.

One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to be more affected by drought. Greater growth reductions for larger trees was first shown on a global scale by @bennett_larger_2015, and numerous subsequent studies have reinforced this finding (*e.g.*, @stovall_tree_2019; @hacket-pain_consistent_2016). It has yet to be resolved which of several potential underlying mechanisms drive this pattern. First, tree height may be a primary driver. Taller trees have a greater biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018]. Vertical gradients in stem and leaf traits--including smaller and thicker leaves (higher LMA), more negative $P50$, and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]--make it biophysically possible for trees to become tall [@couvreur_water_2018]. Meanwhile, tall trees require greater hydraulic efficiency, such that xylem conduit diameter increases with tree height within and across species [@olson_plant_2018; @liu_hydraulic_2019], making large trees more vulnerable to embolism during drought [@olson_plant_2018]. Traits conducive to efficient water transport may also lead to poor ability to recover from or re-route water around embolisms [@roskilly_conflicting_2019]. Second, larger trees may have lower drought resistance because they tend to have more exposed crown positions, where they are exposed to higher solar radiation, greater wind speeds, and lower humidity (e.g., @koike_leaf_2001; @kunert_revised_2017). Subcanopy trees tend to fare better specifically due to the benefits of a buffered environment [@pretzsch_drought_2018]. Potentially counteracting the biophysical challenges faced by large trees, their larger root systems confer a potential advantage in terms of allowing greater access to water; however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure. A final mechanism that could mediate tree size-related responses to drought is how species, and their associated hydraulic traits, are distributed with respect to size [@meakem_role_2018; @liu_hydraulic_2019]. Understanding the mechanisms driving the greater growth reductions of larger trees during drought will require sorting out the interactive effects of height, canopy position, root water access, and species' traits.

Debates have also arisen regarding the traits that would influence tree growth responses to drought. Commonly-measured traits including wood density ($WD$) and leaf mass per area ($LMA$) have been linked to drought responses in some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011] and other forest biomes around the world [@greenwood_tree_2017]. However, in other cases these traits have failed to link to drought tolerance ([Maréchaux et al. 2019](https://besjournals.onlinelibrary.wiley.com/doi/10.1111/1365-2745.13321)), and the direction of response is not always consistent; for instance, higher $WD$ has been associated with greater drought resistance at a global scale [@greenwood_tree_2017] but correlated negatively with tree performance during drought in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011]. Thus, their role may be due to indirect correlations with other traits within life-history strategies [@hoffmann_hydraulic_2011]. Recent work has shown a great potential for hydraulic traits to predict growth and mortality responses. Hydraulic traits including water potentials at which percent loss conductivity surpass a certain threshold ($P50$, $P80$, $P88$) and hydraulic safety margin have enabled prediction of drought performance [@anderegg_hydraulic_2018] but are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics). More rapidly measurable leaf hydraulic traits with direct linkage to plant hydraulic function are emerging with potential to explain greater variation in plant distribution and function than the more commonly-measured $WD$ and $LMA$ [@medeiros_extensive_2019]. These include leaf area shrinkage upon desiccation ($PLA_{dry}$; [@scoffoni_leaf_2014]) and the leaf water potential at turgor loss point ($\pi_{tlp}$), *i.e.,* the water potential at which leaf wilting occurs [@bartlett_correlations_2016]. The abilities of both $PLA_{dry}$ and $\pi_{tlp}$ to explain tree performance under drought remains untested.

>>>>

### Summary
- As climate change is driving increased drought frequency and severity in many forested regions around the world, mechanistic understanding of the factors conferring drought resistance in trees is increasingly important. The dendrochronological record provides a window through which we can understand how tree size and species’ traits shape tree growth responses during droughts.

- We analyzed tree-ring records for twelve species that comprise 97% of the woody productivity of the 25.6-ha ForestGEO plot in an oak-hickory forest of northern Virginia (USA). We sought to determine how tree size, microenvironment characteristics, and species' traits shaped drought responses across the three strongest regional droughts over a 60-year period (1950 - 2009).

@@

- We conclude that hydraulic traits and tree height influence growth responses during drought, and can explain variation in the tree-ring record spanning historical droughts. Thus, these factors can be useful for predicting future drought responses under climate change.

@@

Forests globally play a critical role in climate regulation [@bonan_forests_2008], yet there remains enormous uncertainty as to how the terrestrial carbon (C) sink, which is dominated by forests, will respond to climate change [@friedlingstein_climatecarbon_2006]. An important aspect of this uncertainty lies with physiological responses of trees to drought [@kennedy_implementing_2019]. In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014]; [Dai et al. 2018](https://link.springer.com/article/10.1007/s40641-018-0101-6)), often despite increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015]. Droughts, intensified by climate change, have been affecting forests worldwide [@allen_global_2010], and are expected to continue as one of the most important drivers of forest change in the future [@allen_underestimation_2015]. Understanding and modeling forest responses to drought requires elucidation of how tree size, microenvironment, and species’ traits jointly influence individual-level drought resistance, and the extent to which their influence is consistent across droughts. However, it has proved difficult to resolve factors affecting tree growth during drought with available forest census data, which only rarely captures extreme drought, and with tree-ring records, which rarely represent many species and size classes at a single site and rarely sampled in association with detailed ecological data. The aim of this study was to pair tree-ring and forest plot data to test how tree size and traits shaped growth responses to historical droughts.

One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to be more severely affected by drought. Greater growth reductions for larger trees was first shown on a global scale by @bennett_larger_2015, and subsequent studies have reinforced this finding (*e.g.*, @stovall_tree_2019; @hacket-pain_consistent_2016). It has yet to be resolved which of the potential underlying mechanisms drive this pattern. First, tree height may be a primary driver. Taller trees have a greater biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018]. Vertical gradients in stem and leaf traits--including smaller and thicker leaves (higher leaf mass per area, LMA), more negative water potential inducing 50% loss of hydraulic conductivity ($P50$), and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]--make it biophysically possible for trees to become tall [@couvreur_water_2018]. Meanwhile, tall trees require greater hydraulic efficiency, such that the xylem conduit diameter at the base of the tree increases with tree height within and across species [@olson_plant_2018; @liu_hydraulic_2019], making large trees more vulnerable to embolism during drought [@olson_plant_2018]. Traits conducive to efficient water transport may also lead to poor ability to recover from or re-route water around embolisms [@roskilly_conflicting_2019]. Second, larger trees may have lower drought resistance because they tend to have more exposed crown positions, where they are exposed to higher solar radiation, greater wind speeds, and lower relative humidity (e.g., @koike_leaf_2001; @kunert_revised_2017). Subcanopy trees tend to fare better specifically due to the benefits of a buffered environment [@pretzsch_drought_2018]. Large trees tend to have larger root systems, which potentially counteracts some of the biophysical challenges they face by allowing greater access to water; however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure. A final mechanism that could mediate tree size-related responses to drought is how species, and their associated hydraulic traits, are distributed with respect to size [@meakem_role_2018; @liu_hydraulic_2019]. Understanding the mechanisms driving the greater relative growth reductions of larger trees during drought will require sorting out the interactive effects of height, canopy position, root water access, and species' traits.

Debates have also arisen regarding the traits influencing tree growth responses to drought. Commonly-measured traits including wood density ($WD$) and leaf mass per area ($LMA$) have been linked to drought responses in some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011] and other forest biomes around the world [@greenwood_tree_2017]. However, in other cases these traits have failed to link to drought tolerance ([Maréchaux et al. 2019](https://besjournals.onlinelibrary.wiley.com/doi/10.1111/1365-2745.13321)), and the direction of response is not always consistent; for instance, higher $WD$ has been associated with greater drought resistance at a global scale [@greenwood_tree_2017], but it correlated negatively with tree performance during drought in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011]. Thus, their perceived influence on drought resistance may actually reflect indirect correlations with other traits that more directly drive drought responses [@hoffmann_hydraulic_2011]. Recent work has shown a great potential for hydraulic traits to predict growth and mortality responses. Hydraulic traits including water potentials at which percent loss of conductivity surpass a certain threshold ($P50$, $P80$, $P88$) and hydraulic safety margin have enabled prediction of drought performance [@anderegg_hydraulic_2018] but are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics). More rapidly measurable leaf hydraulic traits with direct linkage to plant hydraulic function are emerging with potential to explain greater variation in plant distribution and function than the more commonly-measured $WD$ and $LMA$ [@medeiros_extensive_2019]. These include leaf area shrinkage upon desiccation ($PLA_{dry}$; [@scoffoni_leaf_2014]) and the leaf water potential at turgor loss point ($\pi_{tlp}$), *i.e.,* the water potential at which leaf wilting occurs [@bartlett_correlations_2016]. The abilities of both $PLA_{dry}$ and $\pi_{tlp}$ to explain tree performance under drought remains untested.
